---
title: "Life Expectancy vs healthcare"
author: "Duc-Quang Nguyen | swissinfo.ch"
date: "10 Oct 2016"
output: 
  html_document:
    toc: true
    toc_depth: 3
    theme: simplex
---

# Data links

* [health expenditure in %GDP](http://data.worldbank.org/indicator/SH.XPD.TOTL.ZS)
* [health expenditure per capita](http://data.worldbank.org/indicator/SH.XPD.PCAP.PP.KD)
  * Total health expenditure is the sum of public and private health expenditures as a ratio of total population. It covers the provision of health services (preventive and curative), family planning activities, nutrition activities, and emergency aid designated for health but does not include provision of water and sanitation. Data are in international dollars converted using 2011 purchasing power parity (PPP) rates. 


## highcharter

* [access extra variable in tooltip](https://github.com/jbkunst/highcharter/issues/54)


```{r setup, include=FALSE}
downloadData <- F

iso2.label <- c("CH", "IT", "FR", "JP", "GB", "ES", "US", "CN", "DE", "TR", "PT", "BR", "RU", "ZA")


translation.file <- "input/healthcare expenditure vs life expectancy - Sheet1.csv"

wb.indicators <- c('SP.DYN.LE00.IN', 'SH.XPD.PCAP.PP.KD', 'SH.XPD.TOTL.ZS')

health.datafile <- "input/healthExpenditure_wb.csv"
le.datafife <- "input/lifeExpectancy_wb.csv"
gdp.datafile <- "input/healthExpenditure_asPercGDP_wb.csv"

library(readr)
library(tidyr)
library(dplyr)
library(magrittr)
library(countrycode)
library(ggplot2)
library(swiMap)
library(swiTheme)
library(forcats)
### Getting data in packages

library(WDI)

library(forcats)

### Interactive 
library(htmltools)
library(shiny)
library(swiRcharts)
library(rCharts)
library(highcharter)
```


```{r get data}
if(downloadData) {
  #helper
  getWBdata <- function(ind.code, ind.name = NULL) {
    if(is.null(ind.name)) { ind.name <- ind.code }
    data.dl <- WDI(
      indicator = ind.code,
      start = 1950,  
      end = 2016, 
      extra = TRUE, 
      cache = NULL
    )
    colnames(data.dl)[3] <- 'value'
    data.dl$indicator <- ind.name
    data.dl %>% select(-capital, -longitude, -latitude, -lending)
  }
  
  ## get health expenditure data from WB http://data.worldbank.org/indicator/SH.XPD.PCAP.PP.KD
  he.data <- getWBdata(wb.indicators[2])
  write_csv(he.data, path = health.datafile)
  
  # get life expectancy data
  le.data <- getWBdata(wb.indicators[1])
  write_csv(le.data, path = le.datafife)  
  
  # get GDP per capita
  gdp.data <- getWBdata(wb.indicators[3])
  write_csv(gdp.data, path = gdp.datafile)   
  
} else {
  txt <- loadTranslation(translation.file)
  he.data <- read_csv(health.datafile)
  le.data <- read_csv(le.datafife)
  gdp.data <- read_csv(gdp.datafile)
}
```

```{r merge & wrangle}

df.all <- right_join(
  he.data %>% rename(he = value) %>% select(-region, -income, -indicator), 
  le.data %>% rename(le = value) %>% select(-region, -income, -indicator)
  )  
df.all <- right_join(
  df.all, 
  gdp.data %>% rename(gdp = value) %>% select(-region, -income, -indicator)
)

df.all <- df.all %>% arrange(year)
# remove NA rows
df.all <- df.all[-which(is.na(df.all$he) & is.na(df.all$le) & is.na(df.all$gdp)), ]

# select only the last year with value for 3 indicators
# http://stackoverflow.com/questions/23340150/using-dplyr-window-functions-to-make-trailing-values-fill-in-na-values
df <- df.all %>% group_by(country, iso2c, iso3c) %>%
  fill(he) %>% fill(le) %>% fill(gdp) %>% 
  summarise(he = last(he), le = last(le), gdp = last(gdp)) %>%
  ungroup()

# remove rows with one or more NA
df <- df[-which(is.na(df$he) | is.na(df$le) | is.na(df$gdp)), ]

# get the world bank regions back
df$region <- gsub(" \\(all income levels\\)", "", le.data[match(df$iso2c, le.data$iso2c), 'region'] %>% unlist())
df <- df %>% filter(region != "Aggregates")

# get the iso2 abbreviations for the world regions
regions <- unique(df$region)
names(regions) <- countryTranslations[match(regions, countryTranslations[,'EN']), "iso2"]
if(any(is.na(names(regions)))) stop()
df$region <- names(regions)[match(df$region, regions)]

# compute the loess as a clean data.frame (broom)
# fit <- loess(le ~ he, data = df, span = 0.9) %>% 
#   augment() %>% 
#   arrange(he)
# 
# 
# fit_nls <- nls(le ~ (he ^ b), start = c(b = 0), data = df) %>%
# augment() %>% 
#   arrange(he)
```


```{r interactive chart}
dd <- df 
dd$label <- ifelse(dd$iso2c %in% iso2.label, dd$iso2c, "")

lang <- 'EN'

for (lang in colnames(txt)) {
  dd$reg <- countryTranslation(dd$region, lang)[,2]
  
  # discard data where no translations are available
  dd <- dd[-which(is.na(countryTranslation(dd$iso2c, lang)[,2])),]
  
  
  dd$tooltip <- paste0(
    '<table cellpadding="1" style="line-height:0.9">',
    '<tr><td><strong>', countryTranslation(dd$iso2c, lang)[,2],'</strong></td></tr>',
    '<tr><td><i><small>', dd$reg, '</i></small></td></tr>',
    '<tr><td>', txt["x.lab",lang], ': <strong>', round(dd$he), '</strong> ', txt["he.tp2",lang], '</td></tr>',
    '<tr><td>', txt["gdp.tp",lang], ": <strong>", round(dd$gdp, 1), '</strong></td></tr>',
    '<tr><td>', txt["y.lab2",lang], ": <strong>", round(dd$le, 1), '</strong> ', txt["le.tp2",lang], '</td></tr>',
    '</table>')
  
  hSeries <- hSeries2(
    data.frame(
      x = dd$he,
      y = dd$le,
      z = dd$gdp,
      series = dd$reg,
      name = dd$label,
      tooltip = dd$tooltip
    ), 
    "series")
  
  bc <- highchart(height = 500) %>%
    hc_chart(type = "bubble", spacing = c(12, 4, 0, 0)) %>%
    hc_add_series_list(hSeries) %>% 
    hc_tooltip(
      formatter = JS("function() { return this.point.tooltip;}"), 
      useHTML = TRUE,
      borderWidth = 2,
      padding = 3
    ) %>%
    hc_legend(
      enabled = T, 
      title = list(text = txt['legend.key', lang]),
      align = 'right',
      verticalAlign = 'bottom',
      x = 10,
      y = -68,
      floating = T
    ) %>% 
    hc_add_theme(hc_theme_swi) %>%
    hc_colors(swi_rpal[-c(2,3,4,6)])
  
  
  bc %<>% hc_yAxis(
    title = list(text = txt['y.lab', lang]), max = 85,
    opposite = ifelse(lang == "AR", TRUE, FALSE)
  ) %>%
    hc_xAxis(
      title = list(text = paste0(txt['x.lab', lang])), 
      labels = list(format =  '{value:,.0f} $'), 
      reversed = ifelse(lang == "AR", TRUE, FALSE)
    ) %>%
    hc_plotOptions(
      bubble = list(
        maxSize = "7%",
        minSize = "0.3%",
        marker = list(lineWidth = 0, fillOpacity = 0.5),
        dataLabels = list(
          enabled = T, 
          allowOverlap = T,
          format = '{point.name}', 
          style = list(
            textShadow = F,
            fontSize = "0.9em",
            verticalAlign = "middle",
            fontWeight = "normal"
          )
        ))
    )
  
  source <- paste0( 
    txt['source', lang], ": ", 
    htmlLink(txt['source2.url', lang], txt['source2.name', lang]), " | ",
    htmlLink("http://www.swissinfo.ch", "swissinfo.ch")
  )
  
  save_html(
    tags$html(
      tags$head(includeHTML(style_swi_highcharter())),
      tags$body(   
        tags$h2(txt['title2', lang]),
        div(class = "descr", txt['subtitle', lang]),
        div(class="graphic", bc),
        div(id = "cite", HTML(source)),
        div(class = "footer", HTML(paste0(txt['footer.tp', lang], " ", txt['footer.c', lang]))),
        HTML(iframeresizer)
      )
    ), file = paste0("02_healthEVsLifeE_bubble_", lang, ".html"), libdir = "js" #, background = "#f2f2f2"
  )     
  
  # move to output
  #file.rename(from = outfile, to = paste0("output/", outfile))
  
  
  # copy js to output -------------> TO DO !!!
  
}

    

```